Software architectures to integrate workflow engines in science gateways

### 2016-05-15 paper.tex
Fixed a few typos and inconsistencies

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), 
PMP (Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), RPPL~\cite{1174106}, SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular
clusters, to execute tasks at a high throughput. Some of
these tools also support advanced features such as provenance
tracking, or  redundancy detection across
analyses to avoid re-computation. A wide array of workflows
have been implemented using these pipeline systems and are now shared across
neuroimaging groups world-wide, which represents a tremendous
opportunity for science gateways to leverage. Domain-specific engines
nicely complement e-Science systems that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

In this paper, a
\emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}), using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipeline systems.
Some workflow engines, 
usually from the e-Science community, 
may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Infrastructure consists of the computing and storage
resources involved in workflow execution, as well as the software
services used to access these resources. Infrastructure can be
composed of computing or file servers, databases, clusters, grids or
clouds. Some workflow engines and science gateways may require specific
characteristics, such as the presence of a shared file system between
the computing nodes, the availability of a global task meta-scheduler,
the presence of a file catalog, etc. In the analysis presented below,
such specific requirements are not discussed. Instead, we consider the
infrastructure as an abstract system that can execute tasks and store
data regardless of the enabling mechanisms.

@@

The interactions involved in the architectures are described below and
labeled (\texttt{a},\texttt{b},\texttt{c}, \ldots) as in Figure~\ref{fig:architectures}.

@@

  the infrastructure or prepared for on-the-fly installation, which

@@

\emph{Robustness of workflow execution} measures the likelihood that the
workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture. 
Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure. More precisely, robustness is determined by the number of
software components and interactions represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}, which are specific to workflow
execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components  (\texttt{R$_1$}): number of specific components
involved in workflow executions. The Science
gateway component, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc., and
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. Data control
  (interaction \texttt{c}), for instance, is \emph{not} specific to
  workflow execution since it is also used by the science gateway to
  transfer user data. 
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

@@

\emph{Scalability} corresponds to the ability of an architecture to
cope with high workloads through the following mechanisms: starting multiple engine instances, distributing engines and task scheduling. 
This is measured by assessing the potential
scalability difficulties due to (partial) absence of a given feature in the architecture.  
Features are evaluated using a 3-level metric: 0 means that
the feature is very easy to enable, 1 means that it can be implemented
but with some difficulty, and 2 means that the feature cannot
 be implemented in a realistic scenario. Three different features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): measures the
  potential for a single gateway instance to use more than one engine
  instance simultaneously. Workflow engines may require important
  amounts of resources when several workflows, or large workflows, are
  executed. At some point, it may be required for the science gateway
  to distribute the load among several engines. \texttt{S$_1$}=0 means
  that
  adding a new engine instance is inherently supported in the
  architecture. In this case, elastic engines can be implemented
  through some kind of auto-scaling mechanism to control the number of
  engine instances in the architecture. \texttt{S$_1$}=1 means that
  multiple engine instances requires specific developments in the
  science gateway or workflow engine. \texttt{S$_1$}=2 means that new
  engine instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): measures the potential of distributing
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 means that distributed engines are enabled by
  the architecture, and \texttt{S$_2$}=1 that they require specific
  developments in the workflow engine. 
\item Task scheduling (\texttt{S$_3$}): measures the complexity
  added by the architecture to task scheduling on
  the infrastructure. Task
  scheduling typically depends more on the implementation of specific
  algorithms in the science gateway, workflow engine and
  infrastructure than on the architecture used to integrate the
  workflow engine in the science gateway. Some architectures, however,
  complicate the task scheduling problem by introducing additional
  software layers or creating tasks with specific
  characteristics. \texttt{S$_3$}=0 means that the architecture does
  not add any additional complexity to the scheduling problem,
  and \texttt{S$_3$}=1 otherwise.
\end{itemize}
These three metrics are summed to obtain a global measure of the scalability
potential of the architecture.

\emph{Specific features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): measures the ability to describe
  meta-workflows from existing workflows. Meta-workflows offer an
  additional level of flexibility to build workflows from reusable
  components. \texttt{O$_1$}=0 means that meta-workflows are
  intrinsically enabled by the architecture (i.e. they can be
  implemented using the components and interactions already in place),
  amd \texttt{O$_1$}=1 means that they may be implemented with some
  development effort.

@@

\paragraph{Specific features} Meta-workflows require specific
developments in the workflow engine or science gateway to allow
workflows to execute other workflows.  (\texttt{O$_1$}=1).
 The
science gateway needs to invoke interaction \texttt{d} to retrieve
dine-grained debugging information from the workflow service
(\texttt{O$_2$}=1).

@@

\paragraph{Scalability} By design, new engine instances only require
new agents, which can be easily automated (\texttt{S$_1$=0}). For
instance, auto-scaling rules can be implemented to start new agents
when the workload in the science gateway exceeds a certain
threshold~\cite{lorido2012auto}. Distributed engines are not available
by default (\texttt{S$_2$=1}) and the architecture does not add any
complexity to the task scheduling problem (\texttt{S$_3$=0}).

@@

\paragraph{Specific features}
Meta-workflows can be implemented, which is one of the main interest
of this architecture (\texttt{O$_1$=0}). Note, however, that the
complexity of the architecture increases with the number of engine
types involved in meta-workflows: for instance, a meta-workflow with
child workflows executed by 2 different engine types would require an
additional workflow service and the corresponding interactions.
Debugging information is accessed through interaction \texttt{d} (invoked twice) (\texttt{O$_2$=1}).

@@

Differences in other specific features should not be neglected. Task encapsulation, nested
workflows (with service invocation) and workflow conversion (with service
invocation) are the only architectures that intrinsically support meta-workflows. Besides,
tight integration is the only architecture that allows
fine-grained debugging. 
The availability of fine-grained debugging information may be
critical to efficient user support.

Table~\ref{table:overall} provides an overall comparison between
architectures, based on the metrics in Table~\ref{table:evaluation}.
Note that this analysis is only meant for illustration purposes, since
it assumes that all criteria have equal weights while real systems
would favor some of them. Overall, task encapsulation performs a bit
better than the other architectures and nested workflows stand a bit
behind for the reasons mentioned previously.
\begin{table*}
\footnotesize
\centering
\begin{tabular}{rcccccc}
                                    & \textbf{Tight}
                                    & \textbf{Service}
                                    & \textbf{Task}
                                    & \textbf{Pool}
                                    & \textbf{Nested}
                                    & \textbf{Conversion} \\
  Integration (global score)    &
                                    \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99E899}{0.22}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99D299}{0.44}
                                    & \cellcolor[HTML]{999999}{1.00}
                                      & \cellcolor[HTML]{99DD99}{0.33}\\
Robustness (global score) &
                                \cellcolor[HTML]{99F099}{0.14}
                                    & \cellcolor[HTML]{99D399}{0.43}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99B699}{0.71}
                                    & \cellcolor[HTML]{999999}{1.00}
                                    & \cellcolor[HTML]{99D399}{0.43}\\
  Extensibility (global score)  & \cellcolor[HTML]{99D299}{0.44}
                                     & \cellcolor[HTML]{99D299}{0.44}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99E899}{0.22}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99E899}{0.22}\\
Scalability (global score)  & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}\\
Specific features (global score) & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99FF99}{0.00}\\
                                    & \cellcolor[HTML]{99DA99}\textbf{1.6}
                                    & \cellcolor[HTML]{99B899}\textbf{2.6}
                                    & \cellcolor[HTML]{99FF99}\textbf{0.5}
                                    & \cellcolor[HTML]{99BF99}\textbf{2.4}
                                    & \cellcolor[HTML]{999999}\textbf{3.5}
                                    & \cellcolor[HTML]{99DE99}\textbf{1.5}\\
\end{tabular}
\caption{Overall evaluation. Brighter colors and lower scores indicate better performance. Scores
  are obtained by summing the normalized global scores ($m'$ values) of
  each criterion in Table~\ref{table:evaluation} and the colors are obtained as in Table~\ref{table:evaluation}. }
\label{table:overall}
\end{table*}

@@

Moreover, all the interactions involved in the architectures were
treated equally, whereas some of them are obviously more complex than
others. For example, interaction \texttt{g} (workflow language
conversion) is clearly more complex than interaction \texttt{d}
(service invocation), and may challenge reliability more strongly. However, without entering into the details of a particular
system, quantification of the robustness or the amount of development
associated with each interaction and software component will hardly be
precise.

@@

% Summary of contributions
We have systematically reviewed architectures used to integrate
workflow engines in science gateways. These architectures are
described in a system-independent framework suitable for comparison,
illustrated on real systems, and evaluated using novel quantitative
metrics that allow simple comparison across architectures. We have
discussed the pros and cons of all the presented architectures based
on these metrics.

% Impact of the work
To the best of our knowledge, our work is the first to systematically
review and compare software architectures used to integrate workflow
engines into science gateways. So far, the literature on science gateways and
workflow engines has focused on the description of particular systems,
or on the presentation of a particular architecture.  Instead, our
analysis abstracts and evaluates architectural patterns independently
from any particular system. It provides general insight to integrate
science gateways and workflow engines, which we hope will be useful
for software architects.

@@

This work has been made possible with the support of the Irving Ludmer
Family Foundation and the Ludmer Centre for Neuroinformatics and
Mental Health. The integration between PSOM and CBRAIN was supported
by a Brain Canada Platform Support Grant, as well as the Canadian
Consortium on Neurodegeneration in Aging (CCNA), through a grant from
the Canadian Institute of Health Research and funding from several
partners.  This work is in the scope of the LABEX PRIMES (ANR-11-
LABX-0063) of Universit\'e de Lyon, within the program
``Investissements d’Avenir'' (ANR-11-IDEX-0007) operated by the French
National Research Agency (ANR). This work also falls into the scope of
the scientific topics of the French National Grid Institute (IdG). The
VIP team thanks the site administrators of the European Grid
Initiative and the GGUS support for their help related to the VIP
platform. The CBRAIN team is grateful for the computing cycles,
storage, and support obtained from Compute Canada
(\url{https://computecanada.ca}) and platform development program from
CANARIE (\url{http://www.canarie.ca/}). We also acknowledge the Dutch
national e-Infrastructure with the support of SURF Cooperative, the
Dutch national program COMMIT/ and the High Performance Computing and
Networking (HPCN) Fund of the University of Amsterdam for their
support to the science gateway actitivies at the AMC.  We are also
grateful to the financial support provided by FP7 E-INFRASTRUCTURE
program for financial support to SCI-BUS, SHIWA and ER-flow
projects.

>>>>

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), 
PMP (Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), RPPL~\cite{1174106}, SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular
clusters, to execute tasks at a high throughput. Some of
these tools also support advanced features such as provenance
tracking or  redundancy detection across
analyses to avoid re-computation. A wide array of workflows
have been implemented using these pipeline systems and are now shared across
neuroimaging groups world-wide, which represents a tremendous
opportunity for science gateways to leverage. Domain-specific engines
nicely complement e-Science systems that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

In this paper, a
\emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}), using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow engines and
domain-specific pipeline systems.
Some workflow engines, 
usually from the e-Science community, 
may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Infrastructure consists of the computing and storage
resources involved in workflow execution, as well as the software
services used to access these resources. Infrastructure can be
composed of computing or file servers, databases, clusters, grids or
clouds. Some workflow engines and science gateways may require specific
characteristics, such as the presence of a shared file system between
the computing nodes, the availability of a global task meta-scheduler, or
the presence of a file catalog. In the analysis presented hereafter,
such specific requirements are not discussed. Instead, we consider the
infrastructure as an abstract system that can execute tasks and store
data regardless of the enabling mechanisms.

@@

The interactions involved in the architectures are described below and
labeled from \texttt{a} to \texttt{g} as in Figure~\ref{fig:architectures}.

@@

  the infrastructure or prepared for on-the-fly deployment, which

@@

\emph{Robustness of workflow execution} measures the likelihood that the
workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture. 
Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure. More precisely, robustness is determined by the number of
software components and interactions represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}, which are specific to workflow
execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components  (\texttt{R$_1$}): number of specific components
involved in workflow executions. The Science
gateway component, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc.
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. Data control
  (interaction \texttt{c}), for instance, is \emph{not} specific to
  workflow execution since it is also used by the science gateway to
  transfer user data. 
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

@@

\emph{Scalability} corresponds to the ability of an architecture to
cope with high workloads through the following mechanisms: starting multiple engine instances, distributing engines and task scheduling. 
This is measured by assessing the potential
scalability difficulties due to (partial) absence of a given feature in the architecture.  
 Three different features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): measures the
  potential for a single gateway instance to use more than one engine
  instance simultaneously. Workflow engines may require important
  amounts of resources when several workflows, or large workflows, are
  executed. At some point, it may be required for the science gateway
  to distribute the load among several engines. \texttt{S$_1$}=0 means
  that
  adding a new engine instance is inherently supported in the
  architecture. In this case, elastic engines can be implemented
  through some kind of auto-scaling mechanism to control the number of
  engine instances in the architecture. \texttt{S$_1$}=1 means that
  multiple engine instances requires specific developments in the
  science gateway or workflow engine. \texttt{S$_1$}=2 means that new
  engine instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): measures the potential of distributing
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 means that distributed engines are enabled by
  the architecture, and \texttt{S$_2$}=1 that they require specific
  developments in the workflow engine. 
\item Task scheduling (\texttt{S$_3$}): measures the complexity
  added by the architecture to task scheduling on
  the infrastructure. Task
  scheduling typically depends more on the implementation of specific
  algorithms in the science gateway, workflow engine and
  infrastructure than on the architecture used to integrate the
  workflow engine in the science gateway. Some architectures, however,
  complicate the task scheduling problem by introducing additional
  software layers or creating tasks with specific
  characteristics. \texttt{S$_3$}=0 means that the architecture does
  not add any additional complexity to the scheduling problem,
  and \texttt{S$_3$}=1 otherwise.
\end{itemize}
These three metrics are summed to obtain a global measure of the scalability
potential of the architecture.

\emph{Specific features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): measures the ability to describe
  meta-workflows from existing workflows. Meta-workflows offer an
  additional level of flexibility to build workflows from reusable
  components. \texttt{O$_1$}=0 means that meta-workflows are
  intrinsically enabled by the architecture (i.e. they can be
  implemented using the components and interactions already in place),
  and \texttt{O$_1$}=1 means that they may be implemented with some
  development effort.

@@

\paragraph{Specific features} Meta-workflows require specific
developments in the workflow engine or science gateway to allow
workflows to execute other workflows.  (\texttt{O$_1$}=1).
 The
science gateway needs to invoke interaction \texttt{d} to retrieve
fine-grained debugging information from the workflow service
(\texttt{O$_2$}=1).

@@

\paragraph{Scalability} By design, new engine instances only require
new agents, which can be easily automated (\texttt{S$_1$=0}). Distributed engines are not available
by default (\texttt{S$_2$=1}) and the architecture does not add any
complexity to the task scheduling problem (\texttt{S$_3$=0}).

@@

\paragraph{Specific features}
Meta-workflows can be implemented, which is one of the main interest
of this architecture (\texttt{O$_1$=0}). Note, however, that the
complexity of the architecture increases with the number of engine
types involved in meta-workflows: for instance, a meta-workflow with
child workflows executed by 2 different types of engines would require an
additional workflow service and the corresponding interactions.
Debugging information is accessed through interaction \texttt{d} (invoked twice) (\texttt{O$_2$=1}).

@@

Differences in other specific features should not be neglected. Task encapsulation, nested
workflows (with service invocation) and workflow conversion (with service
invocation) are the only architectures that intrinsically support meta-workflows. Besides,
tight integration is the only architecture that allows
fine-grained debugging, which may be
critical to efficient user support.

Table~\ref{table:overall} provides an overall comparison between
architectures, based on the metrics in Table~\ref{table:evaluation}.
Note that this analysis is only meant for illustration purposes since
it assumes that all criteria have equal weights while real systems
would favor some of them. Overall, task encapsulation performs a bit
better than the other architectures and nested workflows stand a bit
behind for the reasons mentioned previously.
\begin{table*}
\footnotesize
\centering
\begin{tabular}{rcccccc}
                                    & \textbf{Tight}
                                    & \textbf{Service}
                                    & \textbf{Task}
                                    & \textbf{Pool}
                                    & \textbf{Nested}
                                    & \textbf{Conversion} \\
  Integration (global score)    &
                                    \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99E899}{0.22}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99D299}{0.44}
                                    & \cellcolor[HTML]{999999}{1.00}
                                      & \cellcolor[HTML]{99DD99}{0.33}\\
Robustness (global score) &
                                \cellcolor[HTML]{99F099}{0.14}
                                    & \cellcolor[HTML]{99D399}{0.43}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99B699}{0.71}
                                    & \cellcolor[HTML]{999999}{1.00}
                                    & \cellcolor[HTML]{99D399}{0.43}\\
  Extensibility (global score)  & \cellcolor[HTML]{99D299}{0.44}
                                     & \cellcolor[HTML]{99D299}{0.44}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99E899}{0.22}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99E899}{0.22}\\
Scalability (global score)  & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}\\
Specific features (global score) & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99FF99}{0.00}\\
                                    & \cellcolor[HTML]{99DA99}\textbf{1.6}
                                    & \cellcolor[HTML]{99B899}\textbf{2.6}
                                    & \cellcolor[HTML]{99FF99}\textbf{0.5}
                                    & \cellcolor[HTML]{99BF99}\textbf{2.4}
                                    & \cellcolor[HTML]{999999}\textbf{3.5}
                                    & \cellcolor[HTML]{99DE99}\textbf{1.5}\\
\end{tabular}
\caption{Overall evaluation. Brighter colors and lower scores indicate better performance. Scores
  are obtained by summing the normalized global scores ($m'$ values) of
  each criterion in Table~\ref{table:evaluation} and the colors are obtained as in Table~\ref{table:evaluation}. }
\label{table:overall}
\end{table*}

@@

Moreover, all the interactions involved in the architectures were
treated equally, whereas some of them are obviously more complex than
others. For example, interaction \texttt{g} (workflow language
conversion) is clearly more complex than interaction \texttt{d}
(service invocation). However, without entering into the details of a particular
system, quantification of the robustness or the amount of development
associated with each interaction and software component will hardly be
precise.

@@

% Summary of contributions
We have systematically reviewed architectures used to integrate
workflow engines in science gateways. These architectures were
described in a system-independent framework suitable for comparison,
illustrated on real systems, and evaluated using novel quantitative
metrics that allow simple comparison across architectures. We have
discussed the pros and cons of all the presented architectures based
on these metrics.

% Impact of the work
To the best of our knowledge, our work is the first to systematically
review and compare software architectures  to integrate workflow
engines into science gateways. So far, the literature on science gateways and
workflow engines has focused on the description of particular systems,
or on the presentation of a particular architecture.  Instead, our
analysis abstracts and evaluates architectural patterns independently
from any particular system. It provides general insight to integrate
science gateways and workflow engines, which we hope will be useful
for software architects.

@@

This work has been made possible with the support of the Irving Ludmer
Family Foundation and the Ludmer Centre for Neuroinformatics and
Mental Health. The integration between PSOM and CBRAIN was supported
by a Brain Canada Platform Support Grant, as well as the Canadian
Consortium on Neurodegeneration in Aging (CCNA), through a grant from
the Canadian Institute of Health Research and funding from several
partners.  This work is in the scope of the LABEX PRIMES (ANR-11-
LABX-0063) of Universit\'e de Lyon, within the program
``Investissements d'Avenir'' (ANR-11-IDEX-0007) operated by the French
National Research Agency (ANR). This work also falls into the scope of
the scientific topics of the French National Grid Institute (IdG). The
VIP team thanks the site administrators of the European Grid
Initiative and the GGUS support for their help related to the VIP
platform. The CBRAIN team is grateful for the computing cycles,
storage, and support obtained from Compute Canada
(\url{https://computecanada.ca}) and platform development program from
CANARIE (\url{http://www.canarie.ca/}). We also acknowledge the Dutch
national e-Infrastructure with the support of SURF Cooperative, the
Dutch national program COMMIT/ and the High Performance Computing and
Networking (HPCN) Fund of the University of Amsterdam for their
support to the science gateway actitivies at the AMC.  We are also
grateful to the financial support provided by FP7 E-INFRASTRUCTURE
program for financial support to SCI-BUS, SHIWA and ER-flow
projects.
